Higher-Order unification for free!

### 2024-04-15 paper/main.tex
9

\begin{elpicode}
constraint link-eta {
  rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1) % match
    /  (N2 ~$\triangleright$~ G2 ?- link-eta (uvar X LX2) T2) % remove
    |  (relocate LX1 LX2 T2 T2')             % condition
   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                % new goal
}
\end{elpicode}

@@

Different strategies can be used to unify terms of the object language.
% in the meta langauge. 
The first approach that comes to mind consist in implementing \Uo as a
regular routine in the ML as follows:

@@

\noindent
Opting for this method would result in a suboptimal utilization of the logic
programming engine provided by the meta language, as it degrades indexing by
eliminating all data from rule heads. Additionally,
implementing an unification procedure in the meta
language is likely to be significantly slower compared to the built-in one.

% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
Another possibility is to avoid having application and abstraction nodes
in the syntax tree, and use the ones of meta language, as in the following:

@@

\noindent
However, this encoding has two big limitations. Firstly, in CIC, it is
not always feasible to adopt it due to the meta language's type system being
too limited to accommodate for that one the object language. E.g. CIC
can typecheck variadic functions, see for example~\ref{CPDT chlipala}. \\
Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples.

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work was to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC. Section 6 contains a discussion about the use of the
unification procedure of the meta language in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

\todo{cite isabelle's TC, that are baked in}

>>>>

\begin{elpicode}
constraint link-eta {
  rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1)   % match
    /  (N2 ~$\triangleright$~ G2 ?- link-eta (uvar X LX2) T2)   % remove
    |  (relocate LX1 LX2 T2 T2')               % condition
   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                } % new goal
\end{elpicode}

@@

Different strategies can be used to unify terms of the object language.
% in the meta langauge. 
The first approach that comes to mind consist in implementing \Uo as a
regular routine in the ML, i.e. write rules as follows:

@@

\noindent
Opting for this method would result in a suboptimal utilization of the logic
programming engine provided by the ML, as it degrades indexing by
eliminating all data from rule heads. Additionally,
implementing a unification procedure in the ML is likely to be significantly
slower compared to the built-in one.

Another possibility is to avoid having application and abstraction nodes
in the syntax tree, and use the ones of the ML, as in:

@@

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for CIC due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. CIC can typecheck variadic functions~\cite{cpdt}. \\
Second, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the ML.

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC, and that choice alone makes things arder for us.
Section 6 contains a discussion about the use of the
unification procedure of the ML in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

Another work that is, surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.
